A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-24 imaging_demo/ms.tex
Expand bibtex file and content under section 5
From 63a05b7 (master):
=====================

In epic.bib,

(*) Add references Cornwell et al.(2008),
Thyagarajan et al.(2015a), Thyagarajan et al.(2015b)

In ms.tex,

(*) Add content under section 5

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}}) \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})= \left\langle 
E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t $, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra.
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}),
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

@@

>>>>

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}})= \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) 
\right\rangle_t  \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra,
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}).
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

@@

We now investigate the feasibility of implementing the EPIC imager on current
and future radio telescopes.

We have profiled the core routines of EPIC line-by-line for various ranges of
parameters such as antenna filling fraction, maximum baseline 
length, bandwidth and frequency resolution, integration timescale, etc. for 
HERA antenna layouts which are highly compact.

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the slowest step even for dense HERA 
layouts is found to be the spatial two-dimensional FFT in the imaging stage 
relative to applying the sparse matrix gridding convolution, squaring or 
time-averaging. For instance, even in the dense array layout scenario that 
makes these stages perform the slowest, the gridding convolution, squaring and 
time-averaging take up only $\simeq 40$\%, $\simeq 40$\% and $\simeq 15$\% 
respectively of the time taken by the spatial Fourier transform. With sparser 
arrays they will be even faster.

In the visibility-based imaging, the predominant computational cost is at
the X-engine requiring $N_\textrm{a}^2$ complex multiplications per channel 
per correlator writeout timescale.

In the following discussions, we will assume that the computational cost for
the MOFF imaging is determined by the spatial Fourier transform while  
that for visibility-based imaging comes from $N_\textrm{a}^2$ 
cross-correlations. However, if non-linearities such 
as non-coplanarity of baselines \citep{cor08} and wide-field phenomena like 
the {\it pitchfork} effect \citep{thy15a,thy15b} are to be corrected for, 
the antenna based illumination footprint can start becoming less compact
in the measurement plane and can result in a costlier gridding process.

The number of complex multiplications and additions in the spatial Fourier 
transform implemented via Fast Fourier Transform \citep[FFT;][]{coo65} is 
$\approx \beta N_\textrm{g}\log N_\textrm{g}$ where $N_\textrm{g}$ 
is the number of pixels on the grid and $\beta$ is a constant that depends on 
the implementation of twiddle FFT algorithms \citep{bri74}. In our study, we 
set $\beta=5$, a value much more conservative than was indicated in 
\citet{mor11}. We set the number of complex multiplications in the X-engine 
in visibility based imaging to $N_\textrm{a}^2$.

We consider antenna array layouts described by three quantities fundamental
to radio interferometry, namely, maximum baseline length, number of antennas,
and the size of each antenna.

% Figure~\ref{fig:parameter-space-bll-nant-instruments} shows the
